---
title: "Draft Report"
output: word_document
date: "2022-11-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load pacman package - once only
library(forecast)
library(pacman) #package manager
library(corrplot)
library(leaflet)
library(ggplot2)
library(dplyr)
library(factoextra)
library(cluster)
library(pls)
library(randomForest)
library(Metrics)
library(vip)
library("lubridate")
pacman::p_load(tidyverse)
```
# Introduction
- Project : Investigate the storm strength in the Atlantic
- A Dataset about 30 years of the storm in the Atlantic
- Scientific goal : Understand the factors that make hurricane strong and predict the storm strength

# Load and clean the data
- Example : Wind speed 
- Process all other variables (day, month, year, pressure, latitude, etc.) in the same way.
```{r}
x=1:3700 # Initilize a vector contains the column names
col_names <- paste(x) 
wind_speed <- read.csv("vstore.csv",header=FALSE, col.names = col_names)
dim(wind_speed) #Check dimension of all files
```

```{r,include=FALSE}
days <- read.csv("daystore.csv",header=FALSE, col.names = col_names)
months <- read.csv("monthstore.csv",header=FALSE, col.names = col_names)
years <- read.csv("yearstore.csv",header=FALSE, col.names = c("years"))
hours <- read.csv("hourstore.csv",header=FALSE, col.names = col_names)
lat <- read.csv("latstore.csv",header=FALSE, col.names = col_names)
long <- read.csv("longstore.csv",header=FALSE, col.names = col_names)
pressure <- read.csv("pstore.csv",header=FALSE, col.names = col_names)
#Transpose the pressure matrix
pressure <- data.frame(t(pressure[-1]))
wind_shear_strength <- read.csv("shearstore.csv",header=FALSE, col.names = col_names)
potential_wind_strength <- read.csv("vpstore.csv",header=FALSE, col.names = col_names)
```

```{r, include = FALSE}
#Check dimension
dim(days)
dim(months)
dim(years)
dim(hours)
dim(lat)
dim(long)
dim(pressure)
dim(wind_shear_strength)
dim(potential_wind_strength)
```

```{r, include=FALSE}
#Replace '0' values with NA
hours [hours == 0] <- NA
months [months == 0 ]<- NA
lat [lat ==0 ] <- NA
long [long ==0 ] <- NA
wind_shear_strength [wind_shear_strength==0 ] <- NA
potential_wind_strength[potential_wind_strength==0 ] <- NA
pressure[pressure==0 ] <- NA 
```
# Find the average,maximum, minimum value of each variable
- Replace '0' values with NA : Not affect the mean calculation 
```{r , include = TRUE}
wind_speed[wind_speed ==0 ] <- NA 
colMax <- function(data) sapply(data, max, na.rm = TRUE)
colMin <- function(data) sapply(data, min, na.rm = TRUE)
max_windspeed <- colMax (wind_speed)
min_windspeed <- colMin (wind_speed)
avg_windspeed <- colMeans(wind_speed, na.rm = TRUE)
```
```{r, include = FALSE}
avg_months <- data.frame(round(colMeans(months, na.rm = TRUE),digits = 0))
avg_potential_wind <- colMeans(potential_wind_strength, na.rm = TRUE)
max_pressure <- colMax (pressure)
avg_pressure <- colMeans(pressure, na.rm = TRUE)
avg_shear <- colMeans(wind_shear_strength, na.rm = TRUE)
avg_lat <- colMeans(lat, na.rm = TRUE)
avg_long <- colMeans(long, na.rm = TRUE)
```
# Create new dataframe
```{r, include = TRUE}
hurricane_df <- tibble (col_names,      years,         avg_months, 
                        max_windspeed,  avg_windspeed, min_windspeed,
                        max_pressure,   avg_pressure,  avg_potential_wind, 
                        avg_lat,        avg_long,      avg_shear)
```
# Summary data frame
```{r, echo = FALSE}
colnames(hurricane_df) <- c("storm_ID","years","avg_month", "max_windSpeed","avg_windSpeed", "min_windSpeed","max_Pressure","avg_Pressure","avg_potentialWind","avg_Latitude", "avg_Longitude", "avg_shear")
hurricane_df
```
# Exploratory Data Analysis
- The objective of EDA is:
- Observe the relationships between each variables
- Identify which variable is more important than the others 
- Contribute more in the later stage of research in rapid intensification.
- Choose a good first model, investigate and improve your model.

# Plot clustered correlations using summary df
Wind speed and pressure being the major features of a storm, it is worth to explore their relationship
```{r, fig.cap="Figure 1.Correlation plot between all variables", warning = FALSE}
hurricane_matrix <- data.matrix(hurricane_df[,2:12], rownames.force = NA)
M<-cor(hurricane_matrix)
corrplot::corrplot(M,
        order = "AOE",
        method = "circle", 
        type="upper",  # The correlation matrix can be reordered according to the correlation coefficient. This is important to identify the hidden structure and pattern in the #matrix. 
         tl.srt=45, #text label string rotation 
        )
```
# Correlation graph between 2 variables
```{r,  fig.cap="Figure 2. The relationship between Minimum Wind Speed and Maximum Pressure"}
ggplot(data = hurricane_df, aes(x =min_windSpeed , y =  max_Pressure, color = years)) +
    geom_jitter(alpha = 0.3)+
  labs(y = "Max Pressure" , x = "Min Wind Speed")
```

```{r,echo = FALSE, fig.cap = "Figure 3. The relationship between Average Wind Speed and Average Pressure" }
ggplot(data = hurricane_df, aes(x = avg_windSpeed, y =  avg_Pressure, color = years)) +
    geom_jitter(alpha = 0.3)+
    labs(y = "Average Pressure", x = "Average Wind Speed")
```

```{r,echo = FALSE, fig.cap = "Figure 5. The relationship between Month and Average Wind Speed"}
ggplot(data = hurricane_df, aes(x = as.factor(avg_month), y =  avg_windSpeed, color = years)) +
    geom_jitter(alpha = 0.3)+
    labs(y = "Average Wind Speed", x = "Month")
```




# Storm active time
Assume that if the storms die off, their windspeed is 0
```{r}
active_time = as.numeric(colSums(!is.na(wind_speed)))
active_time_df = as.data.frame(active_time)
hurricane_df["active_time"] = active_time
```

The average of active time of the storms are around 120 hours
```{r, fig.cap = "Figure 10. Box plot of active time"}
active_time_boxplot <- ggplot(hurricane_df, aes(y=active_time)) + 
            geom_boxplot() + 
            theme( axis.text.x=element_blank(),
            axis.ticks.x=element_blank())
active_time_boxplot 
```

# Long lasting storm

```{r, fig.cap = "Figure 12. The relationship between Average  Wind Speed and Average Pressure of Long Lasting Storms" }
long_last_storm <- filter(hurricane_df , active_time > 120)
colors <-c("Long-lasting Storm" = "blue")
ggplot(data = hurricane_df, aes(x = avg_windSpeed, y =  avg_Pressure)) +
    geom_point()+
    geom_point(data = long_last_storm ,aes(avg_windSpeed, y =  avg_Pressure, color = "Long-lasting Storm"))+
    scale_color_manual(values = colors) +
    labs(y = "Average Pressure", x = "Average Wind Speed")
```

Long lasting  storm usually occurs in July to Oct!

```{r,echo = FALSE, , fig.cap = "Figure 13. The relationship between Month and Average Wind Speed of Long Lasting Storms"}
colors <-c("Long-lasting Storm" = "blue")
ggplot(data = hurricane_df, aes(x = as.factor(avg_month), y =  avg_windSpeed)) +
    geom_point()+
    geom_point(data = long_last_storm ,aes(x = as.factor(avg_month), y =  avg_windSpeed , color = "Long-lasting Storm"))+
    scale_color_manual(values = colors) +
    labs(y = "Average Wind Speed", x = "Month" , color = "Legend")
```


# Tracking the numbers of strong stroms every year
Create data frame for each type of storms
```{r}
strong_storm <- filter(hurricane_df , max_windSpeed > 70)
normal_storm <- filter(hurricane_df , max_windSpeed > 50, max_windSpeed < 70)
weak_storm  <- filter(hurricane_df , max_windSpeed < 50)
```


```{r,echo = FALSE, fig.cap = "Figure 15. The relationship between Month and Average Wind Speed of Strong, Normal and Weak Storms"}
colors <- c("Weak Storm" = "black", "Normal Storm" = "blue", "Strong Storm" = "red") 
ggplot() +
    geom_point(data = weak_storm, aes(x = as.factor(avg_month), y =  avg_windSpeed, color = "Weak Storm"))+
    geom_point(data = normal_storm ,aes(x = as.factor(avg_month), y =  avg_windSpeed, color = "Normal Storm"))+
    geom_point(data = strong_storm ,aes(x = as.factor(avg_month), y =  avg_windSpeed, color = "Strong Storm"))+
    scale_color_manual(values = colors) +
    labs(y = "Average Wind Speed", x = "Month", color = "Legend") 
```


# Modelling to understand the data
## Linear Model
# https://www.datacamp.com/tutorial/linear-regression-R
#https://www.scribbr.com/statistics/linear-regression-in-r/
#https://www.machinelearningplus.com/machine-learning/complete-introduction-linear-regression-r/
```{r}
mdl1 <- lm(avg_windSpeed ~ avg_month + max_Pressure + avg_Pressure + avg_potentialWind + avg_Latitude + avg_Longitude + avg_shear + active_time , data = hurricane_df)
summary(mdl1)
```

# Adjust the linear model
```{r, echo = FALSE}
mdl2 <- lm(avg_windSpeed ~ max_Pressure+ avg_Pressure + avg_potentialWind +active_time, data = hurricane_df)
summary(mdl2)
```

# Prepare Train and Test data
```{r}
#https://www.machinelearningplus.com/machine-learning/complete-introduction-linear-regression-r/
# Create Training and Test data -
set.seed(100)  # setting seed to reproduce results of random sampling
train_index <- sample(1:nrow(hurricane_df), 0.8*nrow(hurricane_df))  # row indices for training data
train_data <- hurricane_df[train_index, ]  # model training data
test_data  <- hurricane_df[-train_index, ]   # test data
```

# Build the model on training data
```{r}
lmMod <- lm(avg_windSpeed ~ max_Pressure+ avg_Pressure + avg_potentialWind +active_time, data = train_data)  # build the model
lm_predict <- predict(lmMod, test_data)  # predict the wind speed
#calculate RMSE
rmse_lm <- rmse (lm_predict,test_data$avg_windSpeed )
rmse_lm
```





# Random Forest Regression
#https://hackernoon.com/random-forest-regression-in-r-code-and-interpretation
#https://www.geeksforgeeks.org/random-forest-approach-for-regression-in-r-programming/
#https://www.listendata.com/2014/11/random-forest-with-r.html
```{r, include = FALSE}
mtry <- tuneRF(hurricane_df[-1],hurricane_df$avg_windSpeed, ntreeTry=700,
               stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)
best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1]
print(mtry)
print(best.m)
```
```{r}
set.seed(100)
#Fitting Random Forest model
rf <- randomForest(avg_windSpeed ~ . , data = hurricane_df, mtry=6, ntree=700, importance=TRUE)
rf
```

# Visualize variable importance
```{r, fig.cap = "Figure 19. Variable importance"}
vip(rf) 
```

# Make predictions on a test set and calculate RMSE
```{r}
rf1<- randomForest(avg_windSpeed ~ active_time+ avg_Pressure ,mtry = 2, data = train_data, ntree=700, keep.forest=TRUE, importance=TRUE)
rf1
rf_pred <- predict(rf1, newdata = test_data)
rmse_rf = rmse(rf_pred, test_data$avg_windSpeed)
rmse_rf
```

# Research Question (Predict wind strength)
# Create a dataframe consist of strong storms only
```{r}
strong_storm_ID = list()
for(i in 1:nrow(strong_storm)){
  strong_storm_ID[i] = strong_storm[i,1]
}
strong_wind = wind_speed[,as.numeric(unlist(strong_storm_ID))] # Each column is each storm
strong_hour = hours[,as.numeric(unlist(strong_storm_ID))]
strong_years = years[as.numeric(unlist(strong_storm_ID)),]
strong_day = days[,as.numeric(unlist(strong_storm_ID)),]
strong_month = months[,as.numeric(unlist(strong_storm_ID))]
strong_storm_df = tibble(strong_years,t(strong_wind))
colnames(strong_storm_df) <- c("years","wind_Speed")
```

# Counting the number of strong storms each year
```{r, fig.cap = "Figure 20. Number of strong storms of each year"}
strong_storm_count = strong_storm_df %>% count(years)
plot(strong_storm_count, type="o", lwd=1.5, xlab="Years", ylab="Count", main="Number of strong storms each year") 
```

```{r, fig.cap = "Figure 21. Boxplot of Average Wind Speed of each year of Strong Storms"}
boxplot(avg_windSpeed ~ as.factor(years), ylab = " Average Wind Speed", xlab = "Year", las = 2, data = strong_storm)

```

# Using data of the first storm to do prediction
```{r}
strong_storm_ID = list()
for(i in 1:nrow(strong_storm)){
  strong_storm_ID[i] = strong_storm[i,1]
}
strong_wind = wind_speed[,as.numeric(unlist(strong_storm_ID))] # Each column is each storm
# Store data of the strong storm 1
wind1 <- strong_wind[,3] 
# Find the 1st index where wind speed > 70
head_index <- which.max(wind1>70)
# Find the 1st index where wind speed < 70 after a hurricane forms
tail_index <- which.max( wind1[-1:-head_index] < 70) 

month <- strong_month[,1]
day<- strong_day[,1]
hour <- strong_hour[,1]
year <- strong_years[1]
storm1 <- data.frame(wind1)
colnames(storm1 ) <- c("wind")
#Merging date and time of the wind speed
date <- as.Date(with(storm1, paste(year, month, day,sep="-")), "%Y-%m-%d")
date_time <-  date+hours(hour)
storm1$date_time <- as.character(date_time)
storm1[1:5,]
```

```{r}
#https://www.pluralsight.com/guides/time-series-forecasting-using-r
storm1_train = storm1[1:head_index-1,]
storm1_test = storm1[-c(1:head_index-1),]
storm1_test= storm1_test[complete.cases(storm1_test), ] # Keep only the complete rows
```

```{r}
#Fitting ARIMA model
arima_model <- auto.arima(storm1_train$wind)
#Predicting wind speed
fore_arima = forecast::forecast(arima_model, h=nrow(storm1_test))
df_arima = as.data.frame(fore_arima)
storm1_test$arima = df_arima$`Point Forecast`
#Calculate MAPE
mape(storm1_test$wind, storm1_test$arima) 
storm1_test
```

`


```{r}
colors <- c("Actual value" = "Blue", "Predicted Value" = "Red")
ggplot(storm1_test[1:6,], aes(x= date_time, group =1) ) +
    geom_line(aes(y =  wind, color = "Actual value"),size = 0.8)+
    geom_line(aes(y =  arima, color = "Predicted Value"),size = 0.8)+
    labs(y = "Wind Speed", x = "Date and Time", color= "Legend") + 
    scale_color_manual(values = colors) +
    ylim(60, 90) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))
    
    
```

i=1
wind = c()
month = c()
day = c()
hour = c()
year = c()

for (i in 1:10) {
  print(i)
  speed <- strong_wind[,i]
  wind<-c(wind,speed)
  # Find the 1st index where wind speed > 70
  head_index <- which.max(speed>70)
  tail_index <- which.max(speed[-1:-head_index] < 70) 
  month <- c(month, strong_month[,i])
  day<- c(day,strong_day[,i])
  hour <- c(hour,strong_hour[,i])
  year <- c(year,strong_years[i])
}
  storm <- data.frame(wind) 
  
  colnames(storm ) <- c("wind") 
  #Merging date and time of the wind speed
  date <- as.Date(with(storm, paste(year, month, day,sep="-")), "%Y-%m-%d")
  date_time <- date + hours(hour)
  storm$date_time <- date_time
  # Prepare Train and Test Data Set
  storm_train = storm[1:head_index-1,]
  storm_test = storm[head_index : (head_index+ tail_index),]
  storm_test= storm_test[complete.cases(storm_test), ] # Keep only the complete rows
  
  # Perform model fitting
  arima_model <- auto.arima(storm_train$wind)
  
   #Perform prediction and calculate MAPE

  arima_mdl = forecast::forecast(arima_model, h=nrow(storm_test))
  arima_df = as.data.frame(arima_mdl)
  storm_test$arima = arima_df $`Point Forecast`
```

# For loop to go through all the storms
{r, include = FALSE}
# A list to store mape value
mape_list = list()
i=1
for (i in 1:ncol(strong_wind)) 
{
  
  wind <- strong_wind[, i]

  # Find the 1st index where wind speed > 70
  head_index <- which.max(wind>70)
  tail_index <- which.max( wind[-1:-head_index] < 70
  ) 
  month <- strong_month[,i]
  day<- strong_day[,i]
  hour <- strong_hour[,i]
  year <- strong_years[i]
  storm <- data.frame(wind) 
  colnames(storm ) <- c("wind") 
  #Merging date and time of the wind speed
  date <- as.Date(with(storm, paste(year, month, day,sep="-")), "%Y-%m-%d")
  date_time <- date + hours(hour)
  storm$date_time <- date_time
  # Prepare Train and Test Data Set
  storm_train = storm[1:head_index-1,]
  storm_test = storm[head_index : (head_index+ tail_index),]
  storm_test= storm_test[complete.cases(storm_test), ] # Keep only the complete rows
  
  # Perform model fitting
  arima_model <- auto.arima(storm_train$wind)
  
   #Perform prediction and calculate MAPE

  arima_mdl = forecast::forecast(arima_model, h=nrow(storm_test))
  arima_df = as.data.frame(arima_mdl)
  storm_test$arima = arima_df $`Point Forecast`
  mape_list <- append(mape_list,mape(storm_test$wind, storm_test$arima))
}
mean(as.numeric(mape_list))


